Potential for remotely sensing riverine gas exchange and carbon efflux from SWOT observations

### 2021-05-05 docs/manuscript_v2.Rmd
complete overhaul of repo

---
title: "Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Highlights (3-5 points, 85 characters each w/ spaces)
- The BIKER algorithm can remotely sense riverine gas exchange using the SWOT satellite
- BIKER predicts $CO_2$ fluxes when coupled with in situ gas concentration data
- Predicted $CO_2$ fluxes are similar/marginally worse than when using in-situ methods
- Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
- Accuracy is limited by process-level uncertainties and not by the algorithmic setup

@@

##1 Introduction
The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a]. As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a]. This flux $[M/L^2T]$ is calculated as equation 1 for the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$. Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].

@@

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the reach-averaged flow velocity. This is exasperated in ungauged basins where velocity models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

To circumvent this problem in ungauged basins/at drainage-network-scales, to date researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a] that are trained on extensive hydraulics datasets. This is the approach used in upscaling *V* to continental/global river networks and has proven useful in those settings [e.g. @borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a]. However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977]. Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). A potential alternative to this approach is to directly estimate a river’s hydraulic properties using remote sensing (RS) data.

Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). There are many ways to perform RSQ, from calibrating RS data to local channel hydraulics [e.g. @brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models with both in situ and RS data [e.g. @bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information in their hydraulic/hydrologic models (termed 'ungauged approaches'). Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. **I don't know if introducing SWOT via RSQ is the best idea. Maybe I should just introduce SWOT on its own?**

SWOT will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle. It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. In the context of RSQ, a decade of SWOT work has explored the multi-parameter problem of estimating roughness and bathymetry to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a]. They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. While the raw performance of McFLIs is lower than the more sophisticated approaches mentioned above, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].

McFLIs often employ Bayesian inference. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem as it is with McFLIs [@andreadisConstrainingAssimilationSWOT2020d]. Equinality issues refer to under-constrained systems that can have multiple solutions to the same discharge problem [@garamboisInferenceEffectiveRiver2015].

In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales we develop techniques to ingest SWOT's novel measurements of water surface width and height. Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions:

- Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
- How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

To answer these questions, we develop a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates and associated uncertainties that requires only water surface width and slope and therefore no in situ information about a river’s hydraulic properties. We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3). We also quantify the sensitivity of our algorithm to the expected measurement errors implicit in SWOT measurements. Finally, we compare our algorithm against published methods for estimating carbon dioxide evasion ($FCO_2$) from rivers: we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published gauge-based techniques.

##2 Methods
We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm. To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2). Following these methods, the validation setup is described (section 2.3), as is the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the entire study is included (Figure 1).

![Figure 1. Flowchart of methodology used in this study. ‘Extended Upscaling Model’ is detailed in section 2.1. ‘BIKER’ is detailed in section 2.2 and ‘Flux Estimation’ is detailed in section 2.4. See section 2.3 for the validation setup.](outputs/flowchart.jpg)

###2.1 Developing an RS-able $k_{600}$ upscaling model
@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope. This extended earlier work predicting the reaeration coefficient in a similar manner [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \mathbf{(2)}$$
$$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$

While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged, RS-oriented setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the @ulsethDistinctAirWater2019a model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a]. In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al. (2019) k600 upscaling model. Data from Ulseth et al. (2019).](outputs/k600/theory_plot.jpg)

We combined these five regression models into a ‘rule-based regression model’ (equation 3) for river size *i* and model parameters *a* and *b*. Per Figure 2, there are five possible sets of *a* and *b* parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (slope < 0.05), and high energy rivers 0-10m wide (slope > 0.05).

$$k_{600}=a_i(eD)^{b_i} \mathbf{(3)}$$

This model therefore assigns $a_i$ and $b_i$ a priori using just SWOT measurements of water surface width and slope to define *i*. Another implication of this extension of the @ulsethDistinctAirWater2019a model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.

Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original @ulsethDistinctAirWater2019a model, with the goal of at least replicating their performance. We compared it against the various *eD*-based upscaling models from @raymondScalingGasTransfer2012a that are commonly used in the literature. This was done by training all upscaling functions on 80% of the @ulsethDistinctAirWater2019a data and comparing their performance on the remaining 20% of the data. Because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$. The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.

SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches. Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters that cannot be directly observed from SWOT. We also obtain a ‘change in channel area’ *dA* from the river height and width and assuming a rectangular channel geometry.

In order to use Bayesian inference to solve for $k_{600}$, we need a sampling model/likelihood function that estimates the unobservable statisitcal parameters. Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the previously introduced *dA*). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$$k_{600}=a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(4)}$$

Equation 4 is then reworked into a Bayesian sampling model to have all of the ‘data’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 5, where *x* is the SWOT observables and $\Theta$ are a set of statistical parameters). Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written in equation 5 without that transformation for succinctness. The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaling model itself (equation 4). This will be explained in detail below.

$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{upscale}\right)  \mathbf{(5)}$$

>>>>

---
title: "Potential for remotely sensing riverine gas exchange and carbon efflux from SWOT observations"
author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Highlights (3-5 points, 85 characters each w/ spaces)
- The SWOT satellite should be able to indirectly remotely sense riverine gas exchange velocity
- The BIKER algorithm predicts $CO_2$ fluxes when coupled with in situ gas concentration data
- Predicted BIKER $CO_2$ fluxes compare well with in-situ methods
- Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
- Errors are dominated by process-level uncertainties, not parameter inversion

@@

##1 Introduction
Inland waters transport and transform various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a]. **Too big a jump. What is the journal? ‘constituents’ will need to be defined, as will ‘processes’**As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and thus present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a]. This flux $[M/L^2T]$ is expressed as equation 1 given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$. Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].

@@

To directly measure $k_{600}$ in a given stream, scientists use a range of methods mostly involving floating domes or eddy covariance towers to measure at a point in the river, or using injections of tracer gases to estimate reach-scale gas exchange [again, see @hallGasExchangeStreams2020]. Obiously, these types of of measuremnets are preferred but are infeasible when working at the network-scale across potentially tens of thousands of rivers, and so upscaling functions are used to predict $k_{600}$ from readily available river geomorphology parameters [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope $[L/L]$, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are thereby limited by the quality of the final parameter *V*. Since *V* must be measured in the field or itself estimated by other models/empirical methods, uncertainty in *V* must dominates errors in $k_{600}$ estimates. This is exacerbated in ungauged basins that cover large sections of the earth system, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

To circumvent this problem in ungauged basins and at network-scales, researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between in this case *V* and discharge that are trained on extensive hydraulics datasets [@leopoldHydraulicGeometryStream1953a; @borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a]. However, it has long been established that HG parameters are highly variable across landscapes and from river to river [@gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977]. Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). **Xxx give me a sentence or two on the problems with this is approach here. You bury the problem with the sentence “however, it has long” but then cover it up. The transition doesn’t work without closing this paragraph with a supported (i.e. with citations) statement explicitly defining why we can’t do the global scope HG approach. xxx **

A potential alternative to this ‘global-scope’ HG approach is to directly estimate a river’s hydraulic properties (e.g. V) from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, typically in service of remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). There are many ways to perform RSQ, from calibrating RS data to local channel hydraulics [e.g. @brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models with both in situ and RS data [e.g. @bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information in their hydraulic/hydrologic models (termed 'ungauged approaches'). Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. **CRAIG: might need to add unaguged citations here or rewrite perspective**

SWOT is expected to launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle that will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. A specific subset of these ungauged methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a]. These use basic geomorphic theories and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. McFLIs are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].

While all McFLIs to date have been developed in the context of RSQ, there is no reason their logic cannot be used to estimate $k_{600}$, particularly as McFLIs often employ Bayesian inference for equifinal inverse problems. Equifinality refers to an under-constrained mathematical system that have essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and $K_{600}$ estimation. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While these studies are field based and are not applicable to $k_{600}$ estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem, which matches McFLI logic precisely [@andreadisConstrainingAssimilationSWOT2020d].

In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate $K_{600}$. This would in turn improve our understanding of riverine GHG fluxes in ungauged basins. Therefore, we developed a method to estimate $k_{600}$ solely from SWOT measurements, using hydrologic model data to produce SWOT-like data as SWOT has not yet launched [standard practice in the SWOT community- e,g, @durandIntercomparisonRemoteSensing2016b]. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions:

- Is ungauged RS of the gas exchange velocity ($K_{600}$) possible using soon-to-be-available SWOT river measurements?
- If possible, how will this RS method change estimates of riverine carbon efflux at network scales?

To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{600}$ estimates and its explicit Bayesian uncertainity that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’. We validate BIKER on simulated SWOT data for 22 rivers. We also quantify BIKER's sensitivity to the expected SWOT measurement errors: while SWOT data represent a sea change in inland water monitoring, SWOT is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016]. Finally, we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using BIKER and previously published in situ techniques to show how BIKER’s $k_{600}$ estimates translate into carbon efflux.

##2 Methods
To build BIKER, we first develop a mathematical $k_{600}$ model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we set that model as the core of a McFLI (section 2.2). Following method descriptions, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the entire study is given as Figure 1.

![Figure 1. Flowchart of methodology used in this study. We begin by deriving a mathmatical model for k600 that uses only river width and slope to assign its parameters (section 2.1). Then, we implement this model within a McFLI ('BIKER') to estimate k600 solely from river width and water surface slope (section 2.2). Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4). See section 2.3 for the validation setup.](outputs/flowchart.jpg)

###2.1 Developing a $k_{600}$ model for RS
@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to estimate $k_{600}$ from average flow velocity and channel slope. This extended earlier work predicting the dissolved oxygen reaeration coefficient in a similar manner [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by including measurements from steeper rivers and finding that two distinct $k_{600}$ regimes exist in low and high energy rivers. More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting model parameters are significantly different whether *eD* is high or low. The general form of their $k_{600}$ function is reprinted as equation 2, with the two energy regimes represented by subscript *r*. To date, upscaling studies have assumed uniform flow ($S_0=S_h$) in order to train their parameters across large datasets that only have information on $S_0$.

$$k_{600}=a_r(eD)^{b_r}  \mathbf{(2)}$$

While equation 2 as trained by @ulsethDistinctAirWater2019a dataset achieves the goal of this manuscript and provides a best-to-date predictive accuracy for $k_{600}$, it requires a priori knowledge of *eD* to assign its differential model parameters. Because $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976], we therefore need a priori knowledge of *V* to use equation 2 in a McFLI. So, we must rederive the @ulsethDistinctAirWater2019a $k_{600}$ model such that model parameters are solely assigned using the SWOT observables *W* and $S_h$.

Using the @tsivoglouTracerMeasurementReaeration1976 equation for *eD* for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data) in conjunction with Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 2 as equation 3. In the far-right form of equation 3, we use the Manning’s formulation from @hagemannBAMBayesianAMHGManning2017a which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the change in channel area *dA*). Note that we estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of *W* and $S_h$. Finally, equation 3 is also defined by the parameters $a_i$ and $b_i$, which are described next.

**COLIN ABOUT DA: sensitivity analysis? "We investigated the sensitivity of BIKER to this rectangular assumption and found it negligible (xxx section x.x if appropriate xxx)".**

$$k_{600} = a_i(eD)^{b_i} \approx a_i(gSV)^{b_i} \approx a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(3)}$$

As will be explained in section 2.2, the only observations in the BIKER formulation of equation 3 are *W* and $S_h$. So, we must assign $a_i$ and $b_i$ using just *W* and $S_h$. To do this, we use the basic premise that the equation 2 relationship is different for different river sizes *i*. We use *W* and *S* to define *i* and train different $k_{600}$ functions for rivers with the same *i*. We fit this model to the @ulsethDistinctAirWater2019a data (Figure 2) for five possible sets of $a_i$ and $b_i$ parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (*S* < 0.05), and high energy rivers 0-10m wide (slope > 0.05). We find that narrow rivers (< 10m wide) are best modeled by further breaking them down into low energy (*S* < 0.05) and high energy (*S* > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model as proposed by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a (specifically, normally distributed residuals of the predictions). Another implication of this extension of the @ulsethDistinctAirWater2019a model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.

![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al. (2019) k600 upscaling model. We use the differenial relationship between k600 and eD for different river sizes to define a 'rule-based' regression model. This model assigns differeniable model parameters using just river width and slope, following the form in equation 3. Data from Ulseth et al. (2019).](outputs/k600/theory_plot.jpg)

**these two paragraphs need to make it clear before you get to the end that you believe the relationship to be linear, but you will not limit yourself to raymond and/or ulseth. Rather, you wish to find any number of linear relationships that best fit the k600 data as statistically and hydraulicly defined. THIS MIGHT NEED TO BE A NON-LINEAR FUNCTION. THE OTHER PROBLEM IS THAT ALL WE ARE JUST HIGHLIGHTING THAT ALL UPSCALING RESEARCH HAS BASICALLY ASSUMED UNIFORM FLOW....**

Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original @ulsethDistinctAirWater2019a model, with the goal of at least replicating their performance. We compared it against the various *eD*-based $k_{600}$ models from @raymondScalingGasTransfer2012a that are commonly used in the literature. This validation was done using a classic 80/20 split of the @ulsethDistinctAirWater2019a dataset, where 80% of the data was used to fit these models and 20% was used for independent testing of their predictive performance. This meant that the parameter values were slightly different than those published in both papers, but model structures were identical. This was done because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset and we deemed it fairer to compare model peformances using identical data.

###2.2 Developing BIKER
With a validated equation 3 in hand, we turn next to remotely sensing $k_{600}$. The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s McFLI algorithm for ungauged RSQ via Bayesian inference, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020]. These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data. Below, we translate @hagemannBAMBayesianAMHGManning2017a’s ideas from discharge to $k_{600}$.

SWOT will measure river width and height (and thus water surface slope $S_h$) at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD). Following @hagemannBAMBayesianAMHGManning2017a, the observations needed for Bayesian inference are SWOT measurements of *W* and $S_h$, and all other terms in equation 3 (*n*, $A_0$, and $k_{600}$) are treated as statistical parameters that cannot be directly observed from SWOT.

We therefore rework equation 3 into a Bayesian sampling model to have all of the ‘observations’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 4, where *x* are SWOT observables given $\Theta$, a set of statistical parameters). Note that in this formal likelihood specification both sides of equation 4 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 4 without that transformation for succinctness. The $σ_{k600}^2$ parameter refers to the uncertainty inherent in equation 3's estimates. This will be explained in detail below.

$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{k600}\right)  \mathbf{(4)}$$
